---
title: "Chapter 1 synthetic experiment: recovering covariate coefficients"
format: html
bibliography: references.bib
csl: plos-computational-biology.csl
editor_options:
  chunk_output_type: inline
---

```{r}
#| label: setup
#| include: false

library(dplyr)
library(ggplot2)

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = "..") # Chapter 1 scripts use repo-root paths
```

```{r}
#| label: source
#| include: false

source("synthetic/simulate_recovery.R")
```

## Main question

::: {.callout-note appearance="minimal"}
Does the chapter 1 renewal GAM recover $\beta_1$ without bias and with correct uncertainty? How does this change with and without $s(t)$?
:::

```{mermaid}
flowchart LR
  X["Behavioural covariate X_t"] --> R
  C["Temporal trend c(t)"] --> R
  R["True log R(t) = β₀ + β₁X_t + c(t)"] --> I
  I["Renewal outbreak I_t<br/>with / without NB noise"] --> F
  F["Fit 4 nested GAMs<br/>8-week windows, k ∈ {3, 5, 10}"] --> O
  O["Compare with truth<br/>β̂₁ vs β₁ · log R̂(t) vs log R(t) · CRPS"]
```

Primary:

- **Recovery.** Is $\hat{\beta}_1$ unbiased, and does the reported SE match the empirical SD of $\hat{\beta}_1$?
- **False positives.** When $\beta_1 = 0$, how often does the 95% interval exclude 0?
- **Flexibility of $s(t)$.** How much does $\hat{\beta}_1$ shrink towards 0 as $k$ increases, and does this depend on the form of $c(t)$?

Secondary:

- **$\log R(t)$.** Is the true $\log R(t)$ recovered overall, even when $\beta_1$ is not?
- **Noise and coverage.** Does negative binomial observation noise change how much $s(t)$ absorbs, and which of fixed inputs, median incidence or fixed $\theta$ explains chapter 1's low forecast coverage?
- **Window length.** How do recovery and forecasts change with 4, 8 and 12-week windows?
- **Forecasts.** When $\beta_1 = 0$, does the $X_t$ model still beat the baseline on CRPS?

## Background

Chapter 1 fits a single-stage renewal GAM, with behavioural covariates as linear terms on $\log R(t)$ and $s(t)$ excluded from forecasting, which I chose based on similar deviance explained across models and difficulty extrapolating. Chapter 1 doesn't address what covariates add beyond the temporal variation $R(t)$ needs, and how including $s(t)$ affects coefficient estimates.

## Simulating the truth

### Behavioural covariates $X_t$

We impose three different z-scored behavioural covariate time series, which may interact differently with $s(t)$ terms:

1)  **Piecewise:** Constant in 14-day blocks, so each step in $X_t$ shifts $R(t)$. Simple, but a smooth $s(t)$ may approximate it well.
2)  **Noisier:** Piecewise plus daily noise (SD 0.5, re-scored). Harder for $s(t)$ to mimic, so $\beta_1$ should be recovered.
3)  **True Mobility:** The chapter 1 Google Mobility series[^1], with genuine trends and any concurvity with time.

[^1]: Equal-weighted mean of four out-of-home categories following Nouvellet et al. [@nouvellet2021], but with a trailing rather than centred 7-day mean, and z-scored over the study period.

```{r}
#| label: fig-covariates

# All three in one figure, 1 row, 3 columns
```

### Reproduction number $R(t)$

Simulate true $R(t)$ using $\log R(t) = \beta_0 + \beta_1 X_t + c(t)$, with:

- $\beta_1 \in \{0, 0.25\}$, where 0 gives the false positive rate.
- $\beta_0 = -0.05$. As $X_t$ is z-scored, this is $\log R(t)$ at average study-period behaviour rather than $\log R_0$, and setting it just under 0 means the outbreak neither grows nor dies out.

We may add three forms of $c(t)$ into the true $\log R(t)$ to reflect different residual temporal variation possibilities:

- **None.** $c(t) = 0$, so all variation in $R(t)$ comes from $X_t$.
- **Periodic.** $c(t) = a \sin(2\pi t / P)$ with $a = 0.2$ (±20% on $R(t)$) and $P = 56$ days (one cycle per window). This aims to reflect immunity or variant changes, which $s(t)$ can capture but $X_t$ cannot.
- **Concurvity.** $c(t) = a \tilde{X}_t$, a smoothed $X_t$. Part of $R(t)$ then moves with a smooth trend in $X_t$, which both $X_t$ and $s(t)$ can explain, so the GAM cannot cleanly split it between them. This should affect $\hat{\beta}_1$.

### Simulating an outbreak

Simulate using the renewal formula with and without negative binomial observation noise:

- **Deterministic.** $I_t = R(t) \Lambda_t$.
- **Noisy.** $I_t \sim \text{NegBin}(R(t) \Lambda_t, \theta)$ with $\theta = 10$. The SD is about $\mu/\sqrt{\theta}$, so daily counts vary by about 30%.

## Models compared

Four nested models, fitted by REML over chapter 1's 8-week windows, with $k \in \{3, 5, 10\}$ in `s(t, k = k)`. Window length matters as how much $s(t)$ absorbs depends on $k$ relative to it.

- Baseline: intercept + offset
- $+X_t$: behavioural covariate
- $+s(t)$: smooth term, without $X_t$
- $+X_t + s(t)$: both

### Sampling $\hat{\beta}_1$ over many noisy outbreaks

- **Design.** For each setting, simulate a burn-in (so $\Lambda_t$ has a full generation interval) then one 8-week window, fit the four models once, and repeat over 100 NB noise draws.
- **Outputs.** Bias, empirical SD vs mean reported SE, coverage and the false positive rate at $\beta_1 = 0$.
- **Deterministic outbreaks.** One draw is sufficient, and shows whether $\beta_1$ is identifiable with no noise.

### Rolling origin: chapter 1's figures with a known truth

Explores chapter 1's framework when the truth is known, since where the training window sits (pandemic period, behaviour during that window) matters.

- **Design.** Simulate over the real study period with the real mobility series, then fit and forecast at chapter 1's 27 weekly origins with $k = 5$.
- **Outputs.** With a constant true $\beta_1$, plotting $\hat{\beta}_1$ against origin shows how far estimates deviate from estimation alone.
- **Summarising.** Across draws within each origin, then across origins. Neighbouring origins share 7 of 8 weeks, so expect them to be highly correlated.

## Worked example

Example illustration of the Piecewise $X_t$ series with $\beta_1 = 0.25$ and $c(t) = 0$, with no noise.

```{r}
#| label: fig-example

# Three stacked panels:
# True log R(t) and X_t on the same plot
# Simulated incidence curve
# True log R(t) against fitted log R(t), and true beta1 against beta1_hat, for each of the four models
```

## Results

### Coefficient recovery

Compare true $\log R(t)$ against fitted $\log \hat{R}(t)$, and true $\beta_0$ against $\hat{\beta}_0$ where useful, since top-level $R(t)$ matters as much as its individual components.

```{r}
#| label: tbl-recovery
#| eval: false

# recovery is read from the .rds saved by simulate_recovery.R
# sd: how much β̂₁ actually varies across simulated outbreaks with the same true β₁
# mean_se: mgcv reported standard error for β̂₁, averaged across simulated outbreaks
recovery |>
  group_by(covariate_type, beta1, temporal_trend, noise, with_covariate, use_smooth, smooth_k) |>
  summarise(mean_estimate = mean(estimate),
            bias          = mean(estimate - beta1),
            sd            = sd(estimate),
            mean_se       = mean(se),
            coverage      = mean(covers),
            .groups = "drop")

```

```{r}
#| label: fig-beta1-recovery

```

### Flexibility of $s(t)$

```{r}
#| label: fig-beta1-shrinkage

# May compare whether β̂₁ shrinks towards zero as s(t) flexibility increases
```

### Diagnostics

Check each fit converged, and measure concurvity directly from the input covariates.

- `gam.check` for convergence and effective $k$
- Concurvity, full and pairwise, flagging anything \> 0.8
- Residual ACF with and without $s(t)$

```{r}
#| label: tbl-concurvity

```

### Noise and coverage

Compare $\hat{\theta}$ and coverage with and without NegBin noise.

### Forecasts

- Compare CRPS relative to the baseline model.
- Forecast $s(t)$ models by holding $s(t_{\max})$ fixed (see Notes).

```{r}
#| label: tbl-forecast

```

## Relevance to Chapter 1

- Drivers beyond contacts and mobility, such as mask-wearing, contact duration and immunity, change the secondary attack rate, so coefficients need careful interpretation (explains model performance?)
- Whether $s(t)$ stays a diagnostic, showing how much temporal variation can be captured, or is included in fitting and forecasting.
- How adding $s(t)$ affects $\hat{\beta}_1$.
- Whether $\hat{\beta}_1$ is biased when $s(t)$ is excluded, or $X_t$ acts as a proxy for other temporal variation. Can models still recover the behavioural contribution when an additional immunity-like $c(t)$ is present?
- What behavioural covariates add beyond the variation $R(t)$ needs, and what they explain that a smooth term cannot.

## Next steps

- AR(1) or random-walk process noise on $\log R(t)$, as a further form of $c(t)$
- Time-varying $\beta_1(t)$ over rolling windows
- How does in-sample fit compare to forecast performance, with and without $s(t)$? For example, deviance explained against CRPS, or in-sample uncertainty against forecast coverage

## Notes

### Relevant papers

- **Mellor et al. flu admissions** [@mellor2023] - use only smooths of time and forecast by extrapolating the smooth, which assumes a constant growth rate past $t_{\max}$. Forecast skill depends on smooth's flexibility, chosen by out-of-sample score as 1 basis per 5 days over a 9-week training window.
- **Camacho et al. cholera** [@camacho2018] - renewal equation fitted as Poisson regression with log total infectiousness offset. Covariates enter $\log R(t)$ including penalised lagged rainfall term (a DLNM). No time smooth, and report coefficients as fold-changes in $R(t)$.
- **Ramsay et al. concurvity** [@ramsay2003] - when a covariate shares trend with a smooth of time, GAM standard errors are too small. In their simulation the empirical SD of the estimates was 16-23% larger than the reported SE, giving an 18% type I error at the 5% level. Flag concern when the correlation between a covariate and its fit on the other smooth terms exceeds 0.5.

### Points to consider

- **Timing and shape.** Do the timing and shape of behavioural covariates explain something a generic spline cannot? Where covariate trajectories deviate sharply from smooth trends, capturing those noisy or sudden changes may improve forecasts beyond what a smooth trend can.
- **Residual autocorrelation.** How does residual autocorrelation change with and without $s(t)$? If adding $X_t$ removes much of it, behavioural covariates may be explaining temporal trends rather than leaving them to $s(t)$.
- **Wide intervals.** A wide interval around $\hat{\beta}_1$ does not necessarily imply concurvity, as the covariate may barely move in that window.
- **Extrapolating $s(t)$.** A smooth fitted over the training window shouldn't be extrapolated to $s(t_{\max} + h)$. Because of how basis functions behave past the data, this can give extreme (even negative) $s(t)$ values, and hence extreme $R(t)$, which propagate through the renewal forecast. Instead, sample from the posterior of $s(t)$ at $t_{\max}$ and carry it forward, the $\log R(t)$ analogue of Mellor et al.'s constant growth rate.
- **Smoothed incidence.** inc2prev incidence, and so its $R(t)$, is smoothed, so a smooth $s(t)$ may capture $R(t)$ very well by design. If noisy mobility or contact data could explain rapid changes in $R(t)$, that smoothing limits how far we can see the effect of noisy behaviour. In simulation, a noisy $X_t$ gives a noisy $R(t)$, where the smooth-only model should struggle without overfitting while the $X_t$ model captures it, since we know how $R(t)$ relates to behaviour. In chapter 1, behaviour can be noisy but $R(t)$ and incidence are smoothed, so this may need the inc2prev posterior uncertainty rather than individual noisy trajectories.

---

### Notes from 30th September 2026

- Focus on one covariate series and generate one time series of infections. Condition on these two things, and vary the rest.
- What behaviour adds? If you knew exactly what the relationship between contacts and transmission was, you could then explore scenario projections. Perhaps a good outcome from this synthetic experiment would be how reliable or robust those coefficient estimates are. Similar to Ruarai adding school closures and weekends into projections.
- Consider alternatives for extrapolating forwards e.g. random walk which gives variation within trajectories, rather than just between them (currently sample from posterior)
