---
title: "Synthetic checks: how GAM splits signal between s(t) and covariates"
format: html
editor_options:
  chunk_output_type: inline
---

```{r}
#| label: setup
#| include: false

library(mgcv)
library(dplyr)
library(ggplot2)

source("simulate_recovery.R")

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

## What this checks

- Can a known beta1 be recovered, with its confidence interval, and under what conditions?
- How much of that signal does s(t) take, when the truth is known to sit in the covariate?
- Does NB noise change what s(t) absorbs?
- How often does the interval exclude zero when beta1 is truly zero?
- Are the intervals honest, once GI, theta and future covariates are all known?
- Is the covariate identifiable because it is rough, or because it falls outside the span of s(t)?
- How does recovery of beta1 change with window length and k?

## The covariates

Smooth vs rough covariates

```{r}
#| label: fig-covariates

set.seed(1)

covariates <- bind_rows(
  tibble(t = seq_len(synthetic_config$n_days),
         covariate = make_piecewise_covariate(),
         type = "Piecewise"),
  tibble(t = seq_len(synthetic_config$n_days),
         covariate = make_rough_covariate(),
         type = "Rough")
)

ggplot(covariates, aes(x = t, y = covariate)) +
  geom_line() +
  facet_wrap(~ type, ncol = 1) +
  labs(x = "Day", y = "Covariate (z-scored)") +
  theme_minimal()
```

## One simulated outbreak

What the renewal simulation produces, and that Lambda_t comes from realised counts.

```{r}
#| label: fig-outbreak

set.seed(1)

# beta1 is 0 in the config, which would give a flat Rt, so the demo sets its own
demo_beta1 <- 0.3

demo_covariate <- make_piecewise_covariate()
sim <- simulate_outbreak(demo_covariate, beta1 = demo_beta1, noise = FALSE)

outbreak <- bind_rows(
  tibble(t = sim$t, value = sim$incidence, panel = "Incidence"),
  tibble(t = sim$t, value = sim$rt,        panel = "Rt"),
  tibble(t = sim$t, value = sim$covariate, panel = "Covariate")
)

ggplot(outbreak, aes(x = t, y = value)) +
  geom_line() +
  geom_hline(data = tibble(panel = "Rt"), aes(yintercept = 1), linetype = "dashed") +
  facet_wrap(~ panel, ncol = 1, scales = "free_y") +
  labs(x = "Day", y = NULL) +
  theme_minimal()
```

```{r}
#| label: fig-rt-recovery

gi_weights <- make_gi_weights()

frame <- sim |>
  mutate(Lambda_t   = compute_lambda(incidence, gi_weights),
         log_Lambda = log(Lambda_t)) |>
  filter(is.finite(log_Lambda))

# predict() carries the offset, so subtracting log_Lambda leaves log Rt
recovered_rt <- function(fit) exp(predict(fit, newdata = frame, type = "link") - frame$log_Lambda)

rt_compare <- bind_rows(
  tibble(t = frame$t, rt = frame$rt, source = "True"),
  tibble(t = frame$t,
         rt = recovered_rt(fit_synthetic_renewal(sim, use_smooth = FALSE)),
         source = "Recovered, no s(t)"),
  tibble(t = frame$t,
         rt = recovered_rt(fit_synthetic_renewal(sim, use_smooth = TRUE)),
         source = "Recovered, with s(t)")
)

ggplot(rt_compare, aes(x = t, y = rt, colour = source)) +
  geom_line() +
  geom_hline(yintercept = 1, linetype = "dashed") +
  labs(x = "Day", y = "Rt", colour = NULL) +
  theme_minimal()
```

## Fit without s(t)

Whether beta1 is recovered

```{r}
#| label: tbl-fit-no-smooth

fit_no_smooth <- fit_synthetic_renewal(sim, use_smooth = FALSE)

summarise_beta1_recovery(fit_no_smooth, demo_beta1)
```

## Fit with s(t)

How beta1 recovery changes when s(t) is included

```{r}
#| label: tbl-fit-smooth

fit_smooth <- fit_synthetic_renewal(sim, use_smooth = TRUE)

summarise_beta1_recovery(fit_smooth, demo_beta1)
```

## With observation noise

Whether adding NB noise changes what s(t) absorbs

```{r}
#| label: tbl-fit-noise

# Repeats the with/without s(t) fits but on the NB-drawn counts
# One table, one row per s(t) setting
```

```{r}
#| label: fig-noisy-outbreaks

# Incidence from several NB draws, over the noiseless curve
```

## Where the signal goes

True covariate effect, what the fit attributed, and what s(t) absorbed

```{r}
#| label: fig-log-rt-decomposition

# Plots the true and fitted covariate effects alongside fitted s(t)
# All on the log Rt scale
```

## Repeated draws

One draw gives one estimate, many give bias and coverage

```{r}
#| label: tbl-recovery
#| cache: true

set.seed(1)

recovery <- replicate_beta1_recovery(n_draws = 100, beta1 = demo_beta1)

recovery |>
  group_by(covariate_type, use_smooth, noise) |>
  summarise(mean_estimate = mean(estimate),
            bias          = mean(estimate) - demo_beta1,
            sd            = sd(estimate),
            mean_se       = mean(se),
            coverage      = mean(covers),
            .groups = "drop")
```

```{r}
#| label: fig-beta1-recovery

recovery_labelled <- recovery |>
  mutate(smooth_label    = if_else(use_smooth, "With s(t)", "No s(t)"),
         noise_label     = if_else(noise, "NB noise", "Noiseless"),
         covariate_label = if_else(covariate_type == "piecewise", "Piecewise", "Rough"))

ggplot(recovery_labelled, aes(x = smooth_label, y = estimate)) +
  geom_violin() +
  geom_hline(yintercept = demo_beta1, linetype = "dashed") +
  facet_grid(covariate_label ~ noise_label) +
  labs(x = NULL, y = "Recovered beta1") +
  theme_minimal()
```

## Concurvity

How much of the covariate lies inside the span of s(t)

```{r}
#| label: tbl-concurvity

# Concurvity depends on the covariate and the spline basis
# Same value for every draw, so compute it once per covariate
# Reports worst, observed and estimate
```

## Window length and k

Whether recovery and coverage change with window length and k

```{r}
#| label: tbl-window-k

# Sweeps window length and k, one row per combination
```

## Implications for Chapter 1

Discuss whether s(t) stays as diagnostic-only, and whether reported coefficients are affected by including it or not Should s(t) be included for the forecast models, not just for the in-sample fit? What effect would this have on forecast performance, and how would it affect the question we're trying to answer on what each dataset explains?

## Next steps

- Real UK mobility in place of the synthetic covariate
- Process noise on log Rt, as an AR(1) or random walk rather than iid Gaussian
- Time-varying beta1 over rolling windows, matching the rolling-origin design in Chapter 1
- Smoothed covariates, which behave like a smooth and may worsen concurvity
